---
title: "COVID19_Lineage_assignment_changes_through_time_report"
author: "Julie Chih-yu Chen"
date: "November 3, 2021"
output: html_document
---

 The script in this markdown is used to generate the Sankey visualization of changes in Pangolin lineage assignment through time, and report top changes in counts and proportions of lineages.


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = outputDir)
require(googleVis)
require(ggplot2)
require(ggrepel)
require(dplyr)
```


### loading pangolin prediction output: 
lineage_report.csv in each of the folder under "inputDir"
The pangolin, pangolearn and pango (designation) versions are printed

```{r loaddata, echo=FALSE}

pdates <- timev[selectT]
pdatesclean <- timevclean[selectT]
if(sum(duplicated(pdatesclean))>1){
  warning("There are data run on the same date! Please keep only one run per day to avoid an error")
}
outname<-paste0(typeselect,"_",tail(pdatesclean,n=1)) ## using the last date/version as the name to be saved
#print(outname)

###############
### Assignment extraction: only getting sample ID and assigned lineage, remove those that failed qc, take stock of the versions too
###############

linlist<-lapply(pdates,function(pdate){
  datDir<-file.path(inputDir,pdate)
  oplearn=NULL
  
  if(file.exists(file.path(datDir, "lineage_report.csv"))){
    oplearn<-read.csv(file.path(datDir, "lineage_report.csv"))
    print(paste(pdate, "pangolin:",oplearn$pangolin_version[1],"pangoLEARN:",oplearn$pangoLEARN_version[1], "pango:",oplearn$pango_version[1]))
  
    ### remove those that didn't pass qc
    oplearn<-oplearn[oplearn$status=="passed_qc",c("taxon","lineage")]
  }else{ ## gives a warning 
    warning(paste(file.path(datDir, "lineage_report.csv"), "isn't available."))}
  oplearn

})
names(linlist)=pdates


```

```{r cntsLinTime, echo=FALSE}
### the unique lineages in runs selected
alllin <- sort(unique(unlist(lapply(linlist,function(y){y[,2]}))))
print(paste("Total unique number of lineages in these run:",length(alllin)))

### Counts of the lineages in each run
alllincnt <- do.call(cbind,lapply(linlist,function(y){table(factor(y[,2], levels=alllin))}))
colnames(alllincnt)=pdatesclean

```


```{r FCN_calbeforeGvis, include=FALSE}

### keeping all samples that have a change at any stages, including new samples
## options: "changesOnly" or "changesAndAppeared"
## with options="changesOnly", samples that appeard in the middle timepoint and never change to the end won't be displayed, nor the ones that appeared in the last timepoint
## With options = "changesAndAppeared", plotting samples with changes and also samples that appeared in later stage(s) without a requirement for assignment changes 
compLinDiffMulti<-function(listInput, options="changesOnly"){
  listInput<-lapply(listInput, function(y)y[!duplicated(y$taxon),]) ## remove duplicated samples in the same time point
  tmp<-table(unlist(lapply(listInput, function(y)y$taxon)))
  if(options=="changesOnly"){
      commonSamp<-names(tmp)[tmp>1]##samples present in at least two stages
  }
  if(options=="changesAndAppeared"){
    commonSamp<-names(tmp)
  }  
  
  tmp2<-do.call(cbind,lapply(listInput, function(y,commonSamp){
    nseen<-setdiff(commonSamp,y$taxon)
    if(length(nseen)>0){
      y=rbind(y,cbind(taxon=nseen,lineage=rep(NA,length(nseen))))
    }
    rownames(y)=y$taxon;
    y[commonSamp,"lineage"]
    }, commonSamp=commonSamp))
  
  if(options=="changesOnly"){
    
    whichDiff<- apply(tmp2,1,function(y){y=na.omit(y);sum(y!=y[length(y)],na.rm=T)>0}) 
  }
  if(options=="changesAndAppeared"){
     whichDiff<- apply(tmp2,1,function(y){sum(y==y[length(y)],na.rm=T)!=length(y)})
  }

  diffMat<-tmp2[whichDiff,]
  rownames(diffMat)=commonSamp[whichDiff]
  diffMat
  
}

#compChoice2=linlist[c(10,13,20)]
#diffMat2<-compLinDiffMulti(compChoice2)


##function credit, works excellent: https://stackoverflow.com/questions/15282580/how-to-generate-a-number-of-most-distinctive-colors-in-r
get_distinct_hues <- function(ncolor,s=0.5,v=0.95,seed=40) {
  golden_ratio_conjugate <- 0.618033988749895
  set.seed(seed)
  h <- runif(1)
  H <- vector("numeric",ncolor)
  for(i in seq_len(ncolor)) {
    h <- (h + golden_ratio_conjugate) %% 1
    H[i] <- h
  }
  hsv(H,s=s,v=v)
}

## Generate a pie chart of the change pairs to get a sense of the proportion
#top topl=5 change pairs  
simplifiedPie<-function(x,topl=5,nameAdd=""){
  if(length(x)>0){
    crit <- order(x, decreasing=T)<=topl
    x2<-x[crit]
    if(sum(!crit)>0){
      x2<-c(x2,sum(x[!crit]))
      names(x2)[length(x2)]="Others"
    }
    pie(x2,main=paste0(nameAdd,": ",length(x)," changes in total"))
  }else{print(paste("No changes in", nameAdd))}
}

```


### Generation of the sankey plot 
*Only lineage changes within any of the selected time points are plotted
*Samples unavailable in previous time points are labeled as NA
```{r googleVizMulti, echo=FALSE}
### run givsSankey
### multi-level https://stackoverflow.com/questions/45510421/how-to-make-a-googlevis-multiple-sankey-from-a-data-frame

compChoice <- linlist[names(linlist)] ##running all selected previously

diffMat<-compLinDiffMulti(compChoice)

### if one is only interested in specific lineages, then only keeping samples that have been assigned to be one of the targeted lineages
if (!is.null(linFocus)){
  diffMat<-diffMat[rowSums(matrix(diffMat%in%linFocus,nrow=nrow(diffMat)))>0,]
  outname=paste0(outname,"_",paste(linFocus,collapse="_"))
}



### adding the column names for time point notification
diffMat2<-diffMat
for(i in 1:ncol(diffMat)){
  diffMat2[,i]=paste(diffMat[,i],pdatesclean[i],sep="_") 
  
}
counts <- rep(1, nrow(diffMat)) 

dat <- data.frame(diffMat2,counts)


    
#Aggregate data into a three column table, (from, to and counts) for sequential pair-wise time points
#an automatted way, samples weren't available previously are labelled as NA
datVis=NULL
for(i in 1:(ncol(dat)-2)){
  from=dat[,i]
  to=dat[,i+1]
  dat0=data.frame(from,to, counts)
  suppressWarnings(tt0<-dat0 %>%
    group_by(from, to) %>%
    summarise(counts = sum(counts) ) %>%
    ungroup())
  
  datVis<- rbind(datVis,tt0) 
  
}    
#print(nrow(datVis))

### customizing color
uniqOrder=unique(as.vector(t(as.matrix(datVis[,1:2]))))
uniqOrderLin<-as.factor(sapply(strsplit(uniqOrder,"_"),function(y)y[1]))
#print(length(levels(uniqOrderLin))) # the number of unique lineages: colors required

nodeColors=get_distinct_hues(length(levels(uniqOrderLin)))[as.numeric(uniqOrderLin)]
nodeColorsCmd= paste0("{ link: { colorMode: 'gradient'}, node: { colors: ['",paste(nodeColors,collapse="','"),"']} }")

### setting height, with a maximum
hei=ifelse(nrow(datVis)/1.5*10 > maxSankeyPx, maxSankeyPx, nrow(datVis)/1.5*10)

## customized colors
p3 <- gvisSankey(datVis, from="From", to="To", weight="counts",options=list(sankey=nodeColorsCmd, width=paste0(ncol(dat)*1.6*100,"px"), height=paste0(hei,"px"))) 

## plotting sankey for visualization in a browswer
plot(p3)

### saving sankey html for distribution 
htmlstring <- paste(c(p3$html$header,paste(p3$html$chart,collapse = ""),p3$html$caption,p3$html$footer),collapse = "\n")
cat(htmlstring, file=paste0("sankey_",outname,".html"))

```


### Getting the overall counts and percentages of reassignment, consecutive time point comparison
```{r statsOverall, echo=FALSE}
### just get the overall counts and percentage of reassignment, consecutive comparison
totSamp<-sapply(linlist,nrow)
percReassign=NULL
for (i in 1:(length(totSamp)-1)){
  ttmp<-diffMat[,c(i,i+1)]
  print(sum(ttmp[,1] != ttmp[,2],na.rm=T))
  percReassign[i]=sum(ttmp[,1] != ttmp[,2],na.rm=T)/totSamp[i]*100
}
names(percReassign)=paste0(colnames(alllincnt)[-ncol(alllincnt)], "_to_", colnames(alllincnt)[-1])

#the overall percentages of reassignment, consecutive comparison
#print(percReassign)


ggplot(data=data.frame(datesCompared=names(percReassign), percent=percReassign),aes(x=datesCompared, y=percent)) + geom_bar(stat="identity")+ theme(axis.text.x = element_text(angle=45, hjust=1)) + ggtitle(paste0("% samples w a change in assignment: Consecutive comparisons",pdatesclean[length(pdatesclean)]))+ylab("% change in assignment")+xlab("Comparison between consecutive runs")

```

### Examining changes per lineage with counts versus percentage
One way to flag lineages with less steady assignment in question
Figures labelling lineages with more than 50 samples & greater than 10% in changes

```{r statsByLineage, echo=FALSE}
## from diffMat get counts of difference per lineage at any stage vs the next
## this is just the subset of lineage with change
linchangeCnt<-do.call(rbind,lapply(alllin, function(lin0){
  diffMatYNlin<-diffMat==lin0
  colSums(diffMatYNlin, na.rm=T)
  ret=c()
  totCnt=alllincnt[lin0,]
  for (i in 1:(ncol(diffMat)-1)){
    if(totCnt[i]>0){
      ret[i]=sum(diffMatYNlin[,i]==T&diffMatYNlin[,i+1]==F,na.rm=T)
    }else{
      ret[i]=NA
    }
  }
  ret
}))
rownames(linchangeCnt)=alllin
colnames(linchangeCnt)=paste0(colnames(alllincnt)[-ncol(alllincnt)], "_to_", colnames(alllincnt)[-1])



#from diffMat, get proportion of difference per lineage at any stage vs the next
linpercchange<-do.call(rbind,lapply(alllin, function(lin0){
  diffMatYNlin<-diffMat==lin0
  colSums(diffMatYNlin, na.rm=T)
  ret=c()
  totCnt=alllincnt[lin0,]
  for (i in 1:(ncol(diffMat)-1)){
    if(totCnt[i]>0){
      ret[i]=sum(diffMatYNlin[,i]==T&diffMatYNlin[,i+1]==F,na.rm=T)/totCnt[i]*100
    }else{
      ret[i]=NA
    }
  }
  ret
}))
rownames(linpercchange)=alllin
colnames(linpercchange)=paste0(colnames(alllincnt)[-ncol(alllincnt)], "_to_", colnames(alllincnt)[-1])


for(comparei in 1:ncol(linchangeCnt)){
  cntvsperc<-data.frame(count=linchangeCnt[,comparei], percent=linpercchange[,comparei])
  
  #currently arbitrary
  critLab<- cntvsperc$count>50 & cntvsperc$percent >10 
  
  p<-ggplot(cntvsperc,aes(count, percent,label=ifelse(critLab,rownames(cntvsperc),"")) )+geom_point(color = ifelse(critLab,"red","black"))+ scale_x_continuous(trans='log10')
  
  suppressWarnings(print(p + geom_text_repel()+ labs(title = colnames(linchangeCnt)[comparei])))
}

## extracting percent change by lineage of interest across runs
#linpercchange[c("AY.4","B.1.617.2"),]

### changes through time for the latest top 10 lineages, IF there are changes
#linpercchange[names(head(sort(alllincnt[,ncol(alllincnt)], decreasing=T),10)),]

### overall distribution of % changes between consecutive runs
hist(linpercchange, breaks=30, main="Overall Distribution of Percent Changes between consecutive runs", xlab="Percent change",xlim=c(0,100))

## percentile of changes by lineages
#quantile(na.omit(as.vector(linpercchange)),probs=c(seq(0,0.8,0.1),seq(0.9,1,0.01)))

### printing lineages with greater than 90% change
print("printing lineages with greater than 90% change")
(linpercchange.s<-linpercchange[rowSums(linpercchange>90,na.rm=T)>0,])



```

### Examining pairs of lineage changes between two timepoints
```{r grabbing the pairs of changes, echo=FALSE}

changePairs<-lapply(1:(ncol(diffMat)-1),function(i){
  tc<-diffMat[,i]!=diffMat[,i+1]& !is.na(diffMat[,i]) & !is.na(diffMat[,i+1]) 
  sort(table(paste(diffMat[tc,i],diffMat[tc,i+1],sep="_")), decreasing=T)
})
names(changePairs)=colnames(linpercchange)

## print top 10 change pairs or those with changes of > 50 samples 
lapply(changePairs,function(y) head(y,max(10,sum(y>50))))


#Pie chart generation
for(y in 1:length(changePairs)){
  simplifiedPie(changePairs[[y]],nameAdd=names(changePairs)[y])
}




```


